ASHA 2023 Listening Effort Data Analysis

### 2023-10-29 ASHA 2023 Analysis/data_analysis.Rmd
Updated pupil dilation visualization

## Final Model

@@

@@

## Descriptives and Visualization

@@

data %>%
  # For this analysis, we are only looking at accurate repetition trials
  dplyr::filter(rep_acc == "accurate") %>%
  dplyr::group_by(timebins, speaker) %>%
  dplyr::summarize(pupil_dilation = mean(pupil.binned, na.rm = T),
                   pupil_se = sd(pupil.binned, na.rm = T)/sqrt(n())) %>%
  ggplot() +
  aes(x = timebins,
      y = pupil_dilation,
      color = speaker,
      fill = speaker) +
  geom_ribbon(aes(ymin = pupil_dilation - pupil_se, ymax = pupil_dilation + pupil_se), alpha = .3) +
  geom_line() +
  coord_cartesian(xlim = c(0,5000), ylim = c(-100, 200)) +
  labs(x = "Time (ms)", y = "Pupil Dilation (arbitrary units)", fill = "Speaker", color = "Speaker") +
  theme_bw() + 
  theme(legend.position = "bottom", aspect.ratio = 1)

@@

# Fully Unconditional Model

First, we ran an unconditional growth model. I first specified a model with the orthogonal polynomials added as fixed effects and subject (i.e. listener) and code (i.e. phrase) as random effects. I also added the orthogonal polynomials as random slopes. I also used the AR(1) covariance structure to specifiy the random effect structure. I was not able to run the model. Therefore, I attempted to respecify the random effect structure by removing the AR(1) specification. This led to a model with singular fit. In response to this, I decided to remove code (i.e.) from the random effect structure in order to simplify it.

@@

pupil_roi <- data %>%
  # For this analysis, we are only analyzing pupil dilation for accurate responses
  dplyr::filter(rep_acc == "accurate") %>%
  # filtering out timebins below 500 (i.e., keeping timebins above 500 ms)
  dplyr::filter(timebins >= 500) %>%
  dplyr::group_by(subject, trial) %>%
  dplyr::filter(timebins <= max(timebins) - 2000) %>%
  ungroup() %>%
  # For the purpose of modeling, we are centering time at 500 ms (i.e. the beginning ROI time)
  dplyr::mutate(time_c = timebins - 500) %>%
  # Rescaling pupil size 
  dplyr::group_by(subject, trial) %>%
  dplyr::mutate(norm_pupil = as.numeric(scale(pupil.binned))) %>%
  dplyr::ungroup()

# Polynomials

@@

# Exporting region of interest df

>>>>

### Final Model

@@

# Remove unneeded items from the environment

rm(m0, m1, m2, m3, m4, per_effort_des, perceived_effort_df, per_effort_model_final, model_0, model0)

@@

The ALS speaker has a slower speech rate compared to the control speaker. Therefore, we are going to calculate the average trial legnth to see on average, how different the lengths were for ALS trials and control speaker trials. We will need this information to be able to mark the average phrase offset time and average end point for the analysis region of interest for visualization purposes.

@@

data %>%
  # Selecting relevant parameters
  dplyr::select(subject, trial, speaker, timebins, code) %>%
  # Filtering out time before start of phrase
  dplyr::filter(timebins >= 0) %>%
  dplyr::group_by(speaker, code) %>%
  dplyr::summarize(length = max(timebins)) %>%
  dplyr::group_by(speaker) %>%
  dplyr::summarize(av_length = mean(length),
                   av_offset = av_length - 3000,
                   av_end_roi = av_length - 2000) %>%
  gt::gt() %>%
  gt::cols_label(speaker = "Speaker",
                 av_length = "Average Trial Length (ms)",
                 av_offset = "Average Phrase Length (ms)",
                 av_end_roi = "Average Endpoint for ROI (ms)")

@@

Average trial length includes the phrase presentation plus the 3000 ms retention period. Average phrase legnth is the average length of the recording. Average Endpoint of the Analysis ROI is the 1000 ms after the average phrase endpoint (calculated this for data visualization purposes). On average, the phrase length (phrase presentation), was about 2461 ms longer for the ALS trials than the control speaker trials.

### Visualization

This is the smoothed pupil dilation tracking for the Control Speaker. The solid line indicates the start of the phrase. The dotted line indicates the average phrase offset. The grey box represents the region of interest for analysis.

control <- data %>%
  dplyr::filter(rep_acc == "accurate") %>%
  dplyr::filter(speaker == "Control") %>%
  ggplot() +
  aes(x = timebins,
      y = pupil.binned,
      color = speaker) +
  geom_smooth() +
  geom_vline(xintercept = 0, size = 0.35) +
  geom_vline(xintercept = 1908, linetype = 2) +
  annotate("rect", xmin = 500, xmax = 2908, ymin = -50, ymax = 150, alpha = 0.2) +
  coord_cartesian(xlim = c(0,4910), ylim = c(-100, 200)) +
  labs(title = "Control Speaker", x = "Time (ms)", y = "Pupil Dilation (arbitrary units)") +
  theme_bw() +
  scale_color_manual(values = my_pal[1]) +
  theme(legend.position = "none", aspect.ratio = 1)

control

This is the smoothed pupil dilation tracking for the ALS Speaker. The solid line indicates the start of the phrase. The dotted line indicates the average phrase offset. The grey box represents the region of interest for analysis.

als <- data %>%
  dplyr::filter(rep_acc == "accurate") %>%
  dplyr::filter(speaker == "ALS") %>%
  ggplot() +
  aes(x = timebins,
      y = pupil.binned,
      color = speaker) +
  geom_smooth() +
  geom_vline(xintercept = 0, size = 0.35) +
  geom_vline(xintercept = 4369, linetype = 2) +
  annotate("rect", xmin = 500, xmax = 5369, ymin = -50, ymax = 150, alpha = 0.2) +
  coord_cartesian(xlim = c(0,7369), ylim = c(-100, 200)) +
  labs(title = "ALS Speaker", x = "Time (ms)", y = "Pupil Dilation (arbitrary units)") +
  theme_bw() +
  scale_color_manual(values = my_pal[2]) +
  theme(legend.position = "none", aspect.ratio = 1)

als

pupil_plot <- ggarrange(control, als, ncol = 2)

ggsave("Plots/Pupil_dilation.png", plot = pupil_plot, width = 10, height = 5, units = "in")

## Pupil Dilation Analysis

First, we ran an unconditional growth model. In the intial model, I first specified a model with the orthogonal polynomials added as fixed effects and subject (i.e. listener) and code (i.e. phrase) as random intercepts. I also added the orthogonal polynomials as random slopes. I also used the AR(1) covariance structure to specifiy the random effect structure. However, I was not able to run the model and ran into convergence issues.

Therefore, I decided to respecify the random effect structure by removing the AR(1) specification. This led to a model with singular fit. In response to this, I decided to remove code (i.e.phrase) from the random effect structure in order to simplify it.

More details about how I respecified the model can be seen in the code block.

@@

pupil_roi <- data %>%
  # For this analysis, we are only analyzing pupil dilation for accurate responses
  dplyr::filter(rep_acc == "accurate") %>%
  # filtering out timebins below 500 (i.e., keeping timebins above 500 ms)
  dplyr::filter(timebins >= 500) %>%
  dplyr::group_by(subject, trial) %>%
  dplyr::filter(timebins <= max(timebins) - 2000) %>%
  ungroup()

# Adding linear, quadratic, and cubic orthogonal polynomials to the df

@@

# Exporting region of interest df (this can be used to )
